Basics of Hardware Cache Optimization

### 2017-05-21 appendices.tex
Split content into multiple files and tweak stuff

% Preamble {{{1
\documentclass[a4paper]{scrartcl}

\usepackage[utf8]{inputenc} % Assume this file is encoded in UTF-8.
\usepackage[T1]{fontenc}    % Don't fake umlauts etc.
\usepackage{lmodern}        % Use the lmodern font (http://tex.stackexchange.com/a/65103).
\usepackage{microtype}      % Better microtypography (http://www.ctan.org/pkg/microtype)
\usepackage{hyperref}       % Clickable hyperlinks, load before `glossaries`
\usepackage{cleveref}       % Use `\cref{fig:foo}` instead of `figure~\ref{fig:foo}`.
                            % Load after `hyperref`.  See
                            % <https://tex.stackexchange.com/q/36295>
\usepackage{comment}        % Comment out sections of text.
\usepackage{mathtools}      % Improved facilities for typesetting mathematical formulae
\usepackage{dirtytalk}      % ...

% See <https://tex.stackexchange.com/q/10684/78512>.  TODO.
\usepackage{enumitem}
\setlist{noitemsep}

% TODO.  See <https://tex.stackexchange.com/q/119513>.
\crefname{appsec}{Appendix}{Appendices}

% Use ISO 8601, like a reasonable person.  See <https://tex.stackexchange.com/a/152394>.
\usepackage[style=iso]{datetime2}

% TODO: will this just force LaTeX to make worse choices?
% \usepackage[all]{nowidow}

% Source code listings with improved syntax highlighting
\usepackage{minted}
\usemintedstyle{pastie}

% Define a background color for `minted` listings.  See <http://ctan.org/pkg/minted> and
% <https://tex.stackexchange.com/q/150369>.
\usepackage{xcolor}
\definecolor{bg}{rgb}{0.95,0.95,0.95}
\setminted{bgcolor=bg}
% Don't shade the background when using `mintinline`.
\setmintedinline{bgcolor={}}

% ...
\usepackage[titletoc]{appendix}

% Keep using ISO 8601 consistently, like an even more reasonable person.  See
% <https://tex.stackexchange.com/q/231208>.
\usepackage[date=edtf, urldate=edtf, seconds=true]{biblatex}
\addbibresource{paper.bib}

\usepackage[xindy, toc, acronym]{glossaries} % Load after `hyperref`.

\makeglossaries

\usepackage{tikz}
\usetikzlibrary{datavisualization}
% \usetikzlibrary{datavisualization.formats.functions}

% See <https://tex.stackexchange.com/a/155317>, <https://tex.stackexchange.com/a/320521>
% and <https://tex.stackexchange.com/a/75507>.
% \usepackage{tikzscale}

\newacronym{emm}{EMM}{external memory model}
\newacronym{hdd}{HDD}{hard disk drive}
\newacronym{ram}{RAM}{random access machine}
\newacronym{smt}{SMT}{simultaneous multithreading}
\newacronym{tlb}{TLB}{translation lookaside buffer}
\newacronym{ssd}{SSD}{solid-state drive}

\newglossaryentry{l1d}
{
   name=\texttt{L1d},
   description={},
}

\newglossaryentry{l1i}
{
   name=\texttt{L1i},
   description={},
}

\newglossaryentry{l2}
{
   name=\texttt{L2},
   description={},
}

\newglossaryentry{l3}
{
   name=\texttt{L3},
   description={},
}

\newcommand*{\article}{article} % report, paper?

% Define a command that takes exactly 2 arguments, the first one defaulting to `1`.  The
% second argument should be a list of items separated by `,`.  The list item at the
% position specified by the first argument in printed.  See
% <https://tex.stackexchange.com/a/99271>,
% <http://mirrors.ctan.org/macros/generic/listofitems/listofitems-en.pdf>, and
% <https://tex.stackexchange.com/q/276697>.
\usepackage{listofitems}
\newcommand*{\alts}[2][1]{%
   \setsepchar{,}%
   \readlist*\arg{#2}%
   \arg[#1]%
}

% Top matter {{{1
\title{Basics of Hardware Cache Optimization}
% \title{Algorithms for Hardware Caches}
% \title{Basics of CPU Cache Optimization}
% \title{Basics of Optimizing for Hardware Caches}
% \title{Fundamentals of Optimizing for Hardware Caches}
% \title{Introduction to Optimizations for Hardware Caches}
% \title{Optimizing for Hardware Caches}
% \subtitle{The Basics}
\author{Lukas Waymann}

% Body {{{1
\begin{document}
\maketitle

% \newpage

\begin{abstract}
   % What?
   Typical present-day CPUs have two or more levels of caches.  This \article{} presents
   the basic techniques used to optimize program performance based on knowledge about how
   these hardware caches function.

   The abstract \gls{emm} for memory hierarchies is explained and a small selection of
   algorithms \alts{developed for it, analyzed under it} are explored mathematically and
   empirically.
\end{abstract}
\newpage

\tableofcontents
\newpage

\glsresetall % Reset the use status of all acronyms.

\section{Introduction}
% What is this paper about?  Why learn about them?  How much performance is at stake?

% What are hardware caches?  Why caches?
A hardware cache is a \alts{comparatively, relatively} fast and small physical memory.  It
stores a subset of the data present in slower, larger memory that is expected to be used
again soon.  The purpose of this additional memory is to reduce the number of accesses to
the underlying slower storage.

% Hardware caches aren't going away.
There are fundamental reasons that having one single, \alts{uniform, homogeneous} type of
memory is not viable.  No signal can propagate faster than the speed of light.  Thus,
every storage technology can only reach a finite amount of data within a desired access
latency~\cite[2]{afmh}.

The most ubiquitous example for hardware caches \alts{is the hierarchy, are the various
levels (most commonly 2 or 3)} of CPU caches that are found on almost all present-day
CPUs.  They are designated L1 cache, L2 cache, and so on, with L1 being the fastest and
smallest level.  The underlying storage for CPU caches is the main memory.

There are more storage levels that \alts{comprise, constitute} the \emph{memory hierarchy}
of a computer along with CPU caches and main memory.  For example \glspl{hdd} and
\glspl{ssd}.
% Also: registers, internal buffers of HDDs and SSDs, (tapes), ...
% Focus on CPU caches.  Why?
However, swapping to \glspl{hdd} and \glspl{ssd} continues to become somewhat less common
as main memory sizes increase.  Even non-server systems can currently support 64 GiB of
main memory, eliminating the need for swapping to disk under many workloads.

I will focus on how to use CPU caches effectively and the \alts{enabled, resulting}
performance gains in this \article{}.

% TODO: what about TLB?

\section{Motivation} % So what?
% > Three things Really Matter for performance.  The first one is Algorithm, the second
% > one is your code being Non-Blocking, and the third one is Data Locality."
%      -- http://ithare.com/c-performance-common-wisdoms-and-common-wisdoms/

Hardware caches are managed by hardware directly.  They are generally opaque to the
operating system and other programs.  That is, software has no direct control over the
contents of a hardware cache.

% So what?  Why learn about hardware caches?  How much performance do I gain/lose
% depending on how cache-friendly my algorithm/code is?
\alts{Despite this, We will see that despite this}, two algorithms solving the same
problem with the same asymptotic complexity (in the same \(\Theta(g(n))\)) may differ in
performance by two orders of magnitude because of different \emph{memory access
patterns} (\cref{sec:map})~\cite{bigos}.  We will see an example of this in
\cref{sec:vvl}.

In a nutshell, hardware caches are ubiquitous but the performance improvements they
provide are conditional.
\begin{comment}
   To use them effectively,
   % To obtain optimal performance,
   algorithms must be designed and implemented with the architecture
   % design, structure, manner of functioning, inner workings, properties
   of hardware caches in mind.
\end{comment}
Effective use of hardware caches requires knowledge about \alts{how they work, their
architecture}.  Algorithms must be designed and implemented observing \alts{this
knowledge, their interactions with hardware caches}.

% How?

\section{Types of CPU Caches}
% > Most modern desktop and server CPUs have at least three independent caches: an
% > instruction cache to speed up executable instruction fetch, a data cache to speed up
% > data fetch and store, and a translation lookaside buffer (TLB) used to speed up
% > virtual-to-physical address translation for both executable instructions and data.
%      -- https://en.wikipedia.org/wiki/CPU_cache#Overview
% > There are three common types of CPU caches: ...
%      -- Scott Meyers (talk at code::dive)
Current x86 CPUs \alts{generally, typically, commonly} have three main types of caches:
data caches, instruction caches, and \glspl{tlb}%
~\cite[\href{https://youtu.be/WDIkqP4JbkE?t=11m07s}{11:07}]{scott-meyers-talk}.
Some caches are used for data as well as instructions and are called \emph{unified}.%
~\cite[20]{drepper2007}.
\alts{{A processor may have multiple caches of each type, which}, {Multiple caches of each
type may be present, and}} are organised into numerical \emph{levels}
\alts{{starting at 1, the smallest and fastest level,},}
based on their size and speed.
% Each added level is bigger and slower than its predecessor.
% The smallest and fastest is level 1.

% TODO?  The reason to have multiple levels...

% > Often there are separate L1 caches for instructions and data
%      -- Algorithms for Memory Hierarchies, page 3
% > Systems nowaeays have at-least two levels of cache
%      -- Algorithms for Memory Hierarchies, page 172
% > [T]he caches from L2 on are unified caches which contain both code and data
%      -- Drepper, p. 31
% > Later Intel models have shared L2 caches for dual-core processors.  For quad-core
% > processors we have to deal with separate L2 caches for each pair of two cores.
%      -- Drepper, p. 35

% Terminology / Nomenclature.
In practice, a \alts{currently, presently} representative%
\footnote{%
   % https://en.wikipedia.org/wiki/Bobcat_(microarchitecture)
   E.g. for AMD Family 14h processors~\cite[30--32]{14h},
   % https://en.wikipedia.org/wiki/List_of_AMD_CPU_microarchitectures
   % https://en.wikipedia.org/wiki/Zen_(microarchitecture)
   % 32 KiB L1d, 64 KiB L1i, 512 KiB L2, 8 to 16 MiB L3
   AMD Zen (17h)~\cite{zen}, and
   % https://en.wikipedia.org/wiki/Kaby_Lake
   % https://en.wikipedia.org/wiki/Skylake_(microarchitecture)
   % 32 KiB L1d, 32 KiB L1i, 256 KiB L2, 2 to 8 MiB L3
   Intel Skylake desktop processors%
   ~\cite[figure 2-1, table 2-4]{skylake}
   % ~\cite[figure 2-1, \pno~2-2, table 2-4, \pno~2-6]{skylake}.
   % ~\cite[{2-2}, {2-6}]{skylake}.
   % <https://en.wikipedia.org/wiki/Bulldozer_(microarchitecture)> is too weird.
}
x86 cache hierarchy consists of:
\begin{itemize}
   % https://en.wikipedia.org/wiki/Cache_hierarchy#Shared_versus_private
   \item Separate level 1 data and instruction caches of 32 to 64 KiB for each core
      (denoted \gls{l1d} and \gls{l1i} by  \textcite[14--15]{drepper2007}).
      % TODO?  Why have a separate instruction cache?
      Machine instructions in \gls{l1i} are already decoded%
      ~\cite[31, 56]{drepper2007}.
      % ~\cite[14, 31, 56]{drepper2007}.
   % \item A level 2 cache for \say{both code and data}~\cite[31]{drepper2007}.
   \item A unified \gls{l2} cache of 256 to 512 KiB for each core.
   \item Often a unified \gls{l3} cache of 2 to 16 MiB shared between all cores.
   \item TODO: Some \glspl{tlb} I guess.
\end{itemize}

% \subsection{Access Times}
% http://ithare.com/infographics-operation-costs-in-cpu-clock-cycles/
% http://www.getitwriteonline.com/archive/040201hyphadj.htm
\alts{Estimates, Order-of-magnitude estimates} of typical access latencies \alts[2]{are as
follows, are given by \textcite{ithare-cycles}.}%
\footnote{%
   Intel~\cite[table 2-4]{skylake},
   \textcites
   % {ithare-paadl}{ithare-wisdoms}
   [\href{https://youtu.be/WDIkqP4JbkE?t=17m52s}{17:52}, slide 18]{scott-meyers-talk}
   [2--3, 171]{afmh}[16, 20--21]{drepper2007} all give comparable numbers for various
   architectures.
   % [\ppno~16, 20--21, fig. 3.10]{drepper2007}
}

\begin{center}
   \begin{tabular}{ r | c c c c }
             & \gls{l1d} & \gls{l2} & \gls{l3} & Main Memory \\ \hline
      Cycles & 3--4      & 10--12   & 30--70   & 100--150
   \end{tabular}
\end{center}
%
% These are taken from~\textcite{ithare-cycles} but comparable numbers are given by

% > [Instruction] cache is much less problematic than the data cache.
%      -- Drepper, p. 31
The biggest target for optimizations is the data cache.  \say{[Instruction] cache is much
less problematic}~\cite[31]{drepper2007} and optimizations for data and instruction cache
tend to improve \gls{tlb} usage as well%
~\cite[\href{https://youtu.be/WDIkqP4JbkE?t=11m53s}{11:53}]{scott-meyers-talk}.

\section{Basic Concepts}
% \section{Basic Principles}
% \section{Key Terms} % That sounds boring.
Some architectural properties of hardware caches lead to important concepts for using them
effectively.

\subsection{Cache Line} % or Cache Block
% https://en.wikipedia.org/wiki/CPU_cache#Cache_entries
%
% > On x86/x64, cache line is 64 bytes for many years now.
%      -- http://ithare.com/c-for-games-performance-allocations-and-data-locality/
% > In early caches these lines were 32 bytes long; nowadays the norm is 64 bytes.
%      -- Drepper, p. 15
% > It is not possible for a cache to hold partial cache lines.
%      -- Drepper, p. 16
\emph{Cache lines} or \emph{cache blocks} are the unit of data transfer between main
memory and cache.  They have a fixed size, which has been \say{64 bytes for many years} on
x86/x64 CPUs~\cites{ithare-paadl}[\href{https://youtu.be/WDIkqP4JbkE?t=21m41s}{21:41}]
{scott-meyers-talk}.%
\footnote{%
% > The original Pentium 4 processor also had an eight-way set associative L2 integrated
% > cache 256 KB in size, with 128-byte cache blocks.
%      -- https://en.wikipedia.org/wiki/CPU_cache#Example
% TODO: what about the block size of main memory?  Should it be the same?
Line sizes aren't \emph{necessarily} \alts{identical, homogenous} among a CPU's caches.
The Intel Pentium 4 processor had an \texttt{L1d} cache with \say{64 bytes per cache
line}~\cite[p.~9]{pentium4} but an \texttt{L2} cache with \say{128 bytes per cache
line}~\cite[p.~11]{pentium4}.}
\begin{comment}
   \multiplefootnoteseparator%
   % See <https://tex.stackexchange.com/a/71015>.
   \footnote{This can also be checked on the command line:
   \mintinline{bash}{cat /proc/cpuinfo | grep cache_alignment}}
\end{comment}
% > It means that as soon as you've accessed any single byte in a cache line, all the
% > other 63 bytes are already in L1 cache
%      -- http://ithare.com/c-for-games-performance-allocations-and-data-locality/
\alts{%
   This means accessing a single uncached 32-bit integer entails loading another 60
   adjacent bytes.,
   {This means when a single byte has to be loaded, another 63 adjacent bytes will be as
   well.},
   {When, for example, accessing a single byte that isn't already cached, another 63
   adjacent bytes will be loaded.}
}
% Even when compiling C++ for 64-bit systems, `int` is typically 32-bit.

My laptop's AMD E-450 CPU is no exception and both of its data caches have 64-byte cache
lines.\footnote{\Cref{app:cpuinfo} explains how to obtain this information.}
% \begin{minted}[gobble=3]{bash}
%    $ getconf LEVEL1_DCACHE_LINESIZE; getconf LEVEL2_CACHE_LINESIZE
%    64
%    64
% \end{minted}
%stopzone
% \footnote{%
%    \mintinline{bash}!$ getconf LEVEL1_DCACHE_LINESIZE; getconf LEVEL2_CACHE_LINESIZE!\\
%    %stopzone
%    \noindent\mintinline{bash}!64!\\
%    \noindent\mintinline{bash}!64!
% }
We can verify this quite easily.  Consider \cref{lst:line-size}.  It loops over an array
with an increment given at compile time as \texttt{STEP} and measures the processor time.
The results for different values of \texttt{STEP} are plotted in \cref{fig:line-size}.
% Starting from a step size of 16, the time roughly halves every time the step size is
% doubled.  For the first 4 step sizes however, it is almost constant.
As expected, the time roughly halves whenever the step size is doubled --- but only from a
step size of 16.  For the first 4 step sizes, it is almost constant.

% The reason why the loops take the same amount of time has to do with memory.  The
% running time of these loops is dominated by the memory accesses to the array, not by the
% integer multiplications.
%    -- https://igoro.com/archive/gallery-of-processor-cache-effects/
This is because the run times are \alts{primarily due to, dominated by} memory accesses.
Up to a step size of 8, every 64-byte line has to be loaded.  At 16, the values we modify
are 128 bytes apart,%
\footnote{16 \texttt{int64\_t} values of 8 bytes each}
so every other cache line is skipped.  At 32, three out of four cache lines are skipped,
and so on~\cite[cf.][example 2]{gallery}.

\begin{listing}
   \inputminted[firstline=27]{c}{line-size/line-size.c}
   \caption{Loop over \mintinline{text}{array} with increment \mintinline{text}{STEP}}
   \label{lst:line-size}
\end{listing}

\begin{figure}
   \centering
   \begin{tikzpicture}
      \datavisualization
      [scientific axes=clean,
       x axis={logarithmic,
               ticks={major={at={1 as \textbf{1}, 2 as \textbf{2}, 4 as \textbf{4},
                                 8 as \textbf{8}, 16, 64, 256, 1024}},
                      minor={at={32, 128, 512}}},
               label={Step size}, length=0.8\textwidth},
       y axis={logarithmic,
               % ticks={major={at={4, 16, 64, 256}}, minor={at={8, 32, 128}}},
               % ticks={major={at={6, 12, 24, 48, 92, 184}}},
               ticks={major={at={5, 10, 20, 40, 80, 160, 320}}},
               label={Processor time (ms)}, length=6cm},
       visualize as scatter,
       scatter={style={mark=*, mark options={scale=.65}}}]
         % See <https://tex.stackexchange.com/q/198323>.
         data [read from file=line-size/line-size.csv, separator=\space];
   \end{tikzpicture}
   \caption{Processor times for running \cref{lst:line-size}}
   \label{fig:line-size}
\end{figure}

\subsection{Prefetching}
% http://ithare.com/c-for-games-performance-allocations-and-data-locality/
% http://ithare.com/c-for-games-performance-allocations-and-data-locality/2/
% https://en.wikipedia.org/wiki/Cache_prefetching
\emph{Cache prefetching} is a technique by which CPUs recognize predictable access
patterns (for instance sequential access to an array)... TODO

\subsection{Associativity}
% https://en.wikipedia.org/wiki/CPU_cache#Associativity

\subsection{Cache Hits and Misses}
% TODO: what happens when we miss?  How bad is it?  What about \gls{smt}?

\subsection{Spatial Locality}
Data is loaded from bigger, slower memory into smaller, faster memory (e.g. from main
memory into the CPU cache) in \emph{blocks}.

Moving down the memory hierarchy, access latencies increase faster than the
\emph{obtainable} bandwidth: \say{[w]e can still achieve large bandwidths by accessing
many close-by bits together [...].  Access to large \emph{blocks} [emphasis added] of
memory is almost as fast as access to a single bit},~\cite[2]{afmh}.

% TODO: move this.
% Consider the program shown in \cref{lst:array-sum}.  It repeatedly loops over an array
% to compute the sum of its elements.  Before exiting, it prints the CPU time spend
% summing the array.

\subsection{Temporal Locality}
TODO~\cite{drepper2007}.

\subsection{Memory Access Pattern}
\label{sec:map}

\section{Example: \texttt{std::vector} vs. \texttt{std::list}}
\label{sec:vvl}

% \section{Cache Replacement Policies} % Eviction Policies/Strategies/Algorithms
% https://en.wikipedia.org/wiki/Cache_replacement_policies

\section{External Memory Model}
The \gls{emm} is a widely used
% TODO: is it?
extension of the \gls{ram} model.

\begin{listing}
   \inputminted[firstline=8]{c}{array-sum/array-sum.c}
   \caption{This is C code}
   \label{lst:array-sum}
\end{listing}

% This program is likely to have excellent spatial locality: the array...

% \begin{figure}[htb]
\begin{figure}
   \centering
   \begin{tikzpicture}
      \datavisualization
      [scientific axes=clean,
       x axis={label={Array size (KiB)}, length=0.8\textwidth},
       y axis={label={Processor time (ms)}, length=6cm},
       visualize as scatter,
       scatter={style={mark=*, mark options={scale=.65}}}]
         % See <https://tex.stackexchange.com/q/198323>.
         data [read from file=array-sum/size-time.csv, separator=\space];
   \end{tikzpicture}
   \caption{This figure took way too long to create}
   \label{fig:array-sum}
\end{figure}

% See \cref{fig:array-sum}.  What's going on here?

\section{Cache-Oblivious Algorithms} % Cache-Oblivious Model

\section{Parallel Computing}

\subsection{Cache Coherence} % Coherence or coherency?
% https://en.wikipedia.org/wiki/Cache_coherence
% Dirty.  Invalid.

\subsection{False Sharing}

\clearpage
% \appendix
\begin{appendices}
% See <https://tex.stackexchange.com/q/119513>.
\crefalias{section}{appsec}

@@

\section{TODO: Measuring Stuff}

@@

\end{appendices}

\clearpage

\printglossary[type=\acronymtype] % Print the list of acronyms.

\printbibliography[heading=bibintoc]

\end{document}

>>>>

@@

% Performance Analysis
% Run Time Analysis
% Timing Methodology
% Profiling Methodology
% Profiling Pitfalls
% Run Time Measuring
% Benchmarking Methodology
% Monitoring
\section{Profiling Methodology}
% https://en.wikipedia.org/wiki/Profiling_(computer_programming)
% https://en.wikipedia.org/wiki/Processor_affinity
% On `/usr/bin/time` for benchmarks: <http://stackoverflow.com/q/9006596>.
% Andrei Alexandrescu - Writing Fast Code: <https://youtu.be/vrfYLlR8X8k>.
Plenty of things can go wrong when measuring \alts{execution, run} times.
\begin{itemize}
   \item CPU frequency scaling and boosting
      % Not a problem when measuring cycles.
   \item Interrupts
      % Taking the minimum helps.
   \item CPU jumping
      % See <https://youtu.be/vrfYLlR8X8k?t=33m34s>.
   \item Context switches
   \item Warmth of cache
      % See <https://youtu.be/vrfYLlR8X8k?t=41m00s>.
   % \item ...
\end{itemize}
Some effects will \alts{add noise to, pollute} the results or render them irreproducible,
others will invalidate them completely.

@@

### 2017-05-21 paper.tex
Split content into multiple files and tweak stuff

@@

\usepackage{color}
\definecolor{bg}{rgb}{0.95,0.95,0.95}
\setminted{bgcolor=bg}

\usepackage{biblatex}
\addbibresource{paper.bib}

\usepackage[xindy, toc, acronym, nomain]{glossaries} % Load after `hyperref`.

\makeglossaries

@@

\newacronym{emm}{EMM}{external memory model}
\newacronym{hdd}{HDD}{hard disk drive}
\newacronym{ram}{RAM}{random access machine}
\newacronym{smt}{SMT}{simultaneous multithreading}
\newacronym{tlb}{TLB}{translation lookaside buffer}
\newacronym{ssd}{SSD}{solid-state drive}

\newcommand{\article}{article} % report, paper?

% Top matter {{{1
\title{Algorithms for Hardware Caches}
\author{Lukas Waymann}

% Body {{{1
\begin{document}
\maketitle
\newpage

@@

\section{Introduction}
% What is this paper about?  Why learn about them?  How much performance is at stake?

% What are hardware caches?
A hardware cache is a comparatively
% relatively
fast and small physical memory.  It stores a subset of the data present in slower, larger
memory that is expected to be used again soon.  The purpose of this additional memory is
to reduce the number of accesses to the underlying slower storage.

% Why caches?
There are fundamental reasons that having one single, uniform
% homogeneous
type of memory is not viable.  No signal can propagate faster than the speed of light.
Thus, every storage technology can only reach a finite amount of data within a desired
access latency \cite[p.~2]{afmh}.

% TODO: move this paragraph out of the introduction.
% Hardware caches are managed by hardware directly.  They are generally opaque to the
% operating system and other programs.  That is, software has no direct control over the
% contents of a hardware cache.

The most ubiquitous example for hardware caches is the hierarchy of
% are the various levels (most commonly 2 or 3) of
CPU caches that are found on almost all present-day CPUs.  They are designated L1 cache,
L2 cache, and so on, with L1 being the fastest and smallest level.  The underlying storage
for CPU caches is the main memory.

There are more storage levels that comprise
% constitute
the \emph{memory hierarchy} of a computer along with CPU caches and main memory.  For
example \glspl{hdd} and \glspl{ssd}. % Also: registers, internal buffers of HDDs and SSDs,
                                     % (tapes?), ...
% Focus on CPU caches.  Why?
However, swapping to \glspl{hdd} and \glspl{ssd} continues to become somewhat less common
as main memory sizes increase.  Even non-server systems can currently support 64 GiB of
main memory, eliminating the need for swapping to disk under many workloads.

I will focus on how to use CPU caches effectively and the resulting
% enabled
performance gains in this \article{}.
% TODO: what about TLB?

\section{Motivation} % So what?
%    Three things Really Matter for performance. The first one is Algorithm, the second
%    one is your code being Non-Blocking, and the third one is Data Locality."
%       -- http://ithare.com/c-performance-common-wisdoms-and-common-wisdoms/

% We will see that
Two algorithms solving the same problem with the same asymptotic complexity (in the same
\(\Theta(g(n))\)) may differ in performance by two orders of magnitude~\cite[p.~2]{afmh}.

% TODO: how and how much can we speed up our programs by considering their interactions
% with CPU caches?

% TODO: move this.
% Consider the program shown in \cref{lst:array-sum}.  It repeatedly loops over an array to
% compute the sum of its elements.  Before exiting, it prints the CPU time spend summing the
% array.

\begin{listing}
   \inputminted[firstline=3]{c}{array-sum/array-sum.c}
   \caption{This is C code}
   \label{lst:array-sum}
\end{listing}

@@

See \cref{fig:array-sum}.  What's going on here?

% \section{Basic Principles}
% \section{Basic Concepts}
\section{Key Terms}
\subsection{Spatial Locality}
Data is loaded from bigger, slower memory into smaller, faster memory (e.g. from main
memory into the CPU cache) in \emph{blocks}.

Moving down the memory hierarchy, access latencies increase faster than the
\emph{obtainable} bandwidth: \say{[w]e can still achieve large bandwidths by accessing
many close-by bits together [...].  Access to large \emph{blocks} [emphasis added] of
memory is almost as fast as access to a single bit},~\cite[p.~2]{afmh}.

\subsection{Temporal Locality}

\subsection{Cache Hit}

\subsection{Cache Miss}
% TODO: what happens when we miss?  How bad is it?  What about \gls{smt}?

\subsection{External Memory Model}
The \gls{emm} is a widely used
% TODO: is it?)
extension of the \gls{ram} model.

>>>>

% See <https://tex.stackexchange.com/q/10684/78512>.  TODO.
\usepackage{enumitem}
\setlist{noitemsep}

% TODO.  See <https://tex.stackexchange.com/q/119513>.
\crefname{appsec}{Appendix}{Appendices}

% Use ISO 8601, like a reasonable person.  See <https://tex.stackexchange.com/a/152394>.
\usepackage[style=iso]{datetime2}

% TODO: will this just force LaTeX to make worse choices?
% \usepackage[all]{nowidow}

@@

% Define a background color for `minted` listings.  See <http://ctan.org/pkg/minted> and
% <https://tex.stackexchange.com/q/150369>.
\usepackage{xcolor}
\definecolor{bg}{rgb}{0.95,0.95,0.95}
\setminted{bgcolor=bg}
% Don't shade the background when using `mintinline`.
\setmintedinline{bgcolor={}}

% ...
\usepackage[titletoc]{appendix}

% Keep using ISO 8601 consistently, like an even more reasonable person.  See
% <https://tex.stackexchange.com/q/231208>.
\usepackage[date=edtf, urldate=edtf, seconds=true]{biblatex}
\addbibresource{paper.bib}

\usepackage[xindy, toc, acronym]{glossaries} % Load after `hyperref`.

\makeglossaries
\loadglsentries{glossary}

@@

\newcommand*{\article}{article} % report, paper?

% Define a command that takes exactly 2 arguments, the first one defaulting to `1`.  The
% second argument should be a list of items separated by `,`.  The list item at the
% position specified by the first argument in printed.  See
% <https://tex.stackexchange.com/a/99271>,
% <http://mirrors.ctan.org/macros/generic/listofitems/listofitems-en.pdf>, and
% <https://tex.stackexchange.com/q/276697>.
\usepackage{listofitems}
\newcommand*{\alts}[2][1]{%
   \setsepchar{,}%
   \readlist*\arg{#2}%
   \arg[#1]%
}

% Top matter {{{1
\title{Basics of Hardware Cache Optimization}
% \title{Algorithms for Hardware Caches}
% \title{Basics of CPU Cache Optimization}
% \title{Basics of Optimizing for Hardware Caches}
% \title{Fundamentals of Optimizing for Hardware Caches}
% \title{Introduction to Optimizations for Hardware Caches}
% \title{Optimizing for Hardware Caches}
% \subtitle{The Basics}
\author{Lukas Waymann}

% Body {{{1
\begin{document}
\maketitle

% \newpage

@@

\input{intro}
\input{motivation}
\input{types}
\input{basic-concepts}

\section{Example: \texttt{std::vector} vs. \texttt{std::list}}
\label{sec:vvl}

% \section{Cache Replacement Policies} % Eviction Policies/Strategies/Algorithms
% https://en.wikipedia.org/wiki/Cache_replacement_policies

\section{External Memory Model}
The \gls{emm} is a widely used
% TODO: is it?
extension of the \gls{ram} model.

\begin{listing}
   \inputminted[firstline=8]{c}{array-sum/array-sum.c}
   \caption{This is C code}
   \label{lst:array-sum}
\end{listing}

@@

% See \cref{fig:array-sum}.  What's going on here?

\section{Cache-Oblivious Algorithms} % Cache-Oblivious Model

\section{Parallel Computing}

\subsection{Cache Coherence} % Coherence or coherency?
% https://en.wikipedia.org/wiki/Cache_coherence
% Dirty.  Invalid.

\subsection{False Sharing}

\clearpage
% \appendix
\begin{appendices}
   % See <https://tex.stackexchange.com/q/119513>.
   \crefalias{section}{appsec}
   \input{appendices}
\end{appendices}
